\LARGE\textbf{Designing Studies and Evaluating Research Results:\\ Type M and Type S Errors for\\ Pearson Correlation Coefficient }

### 2020-05-28 Documents/Paper_main/Paper_main.Rnw
corrige Giulia

Effect size exaggeration is a corollary consequence of low statistical power that might not be evident at first. Indeed, by definition, statistical power focuses on statistical significance, not effect size estimation (i.e., statistical power is the probability to find a statistically significant result if an effect of a certain size actually exists). However, effect size estimation and statistical significance are closely related. This point can be highlighted estimating the Type M (magnitude) and Type S (sign) errors characterizing a study design \parencite{gelmanPowerCalculationsAssessing2014}. Given a study design (i.e., sample size, statistical test directionality, and $\alpha$ level), Type M error indicates the factor by which a statistically significant effect would be, on average, exaggerated. Type S error indicates the probability to find a statistically significant effect in the opposite direction to the one considered plausible. It is evident that both errors are defined starting from a reasoned guess on the plausible magnitude and direction of the effect under study, which is called \emph{plausible effect size} \parencite{gelmanPowerCalculationsAssessing2014}. The analysis that researchers perform to consider Type M and Type S errors in their research practice is called \emph{design analysis}, given the special focus posed into considering the design of a study \parencite{altoeEnhancingStatisticalInference2020, gelmanPowerCalculationsAssessing2014}.

@@

Type M (magnitude) and Type S (sign) errors are not widely known in the psychological research community but their consideration during the research process has the potential to improve current research practices, for example, by increasing the awarness that design choices have on possible studies' results. In a previous work we illustrated Type M and Type S errors using Cohens’\emph{d} as a measure of effect size \parencite{altoeEnhancingStatisticalInference2020}. In this contribution we aim to to further increase familiarity with Type M and Type S errors, considering another common effect size measures in psychology: Pearson correlation coefficient, $\rho$.  In the following paragraphs we discuss what Type M and Type S errors are and how to perform a design analysis using the ad-hoc R functions that we developed.

@@

Pearson correlation coefficient is a standardized effect size measure indicating the strength and the direction of the relationship between two continuous variables \parencite{cohenStatisticalPowerAnalysis1988, ellisEssentialGuideEffect2010}. Even though the correlation coefficient is widely known, we briefly go over its main features using an example. Imagine that we were interested to measure the relationship between anxiety and depression in a population and we plan a study with \emph{n} participants, where, for each participant, we measure the level of anxiety (i.e., variable X) and the level of depression (i.e., variable Y). At the end of the study we will have n pairs of values X and Y. The correlation coefficient helps us answer the questions: how strong is the relationship between anxiety and depression in this population? Is the relationship positive or negative? Correlation ranges from -1 to +1, indicating respectively two extreme scenarios of perfect negative relationship and perfect positive relationship. Since the correlation coefficient is a dimensionless number, it is a signal to noise ratio where the signal is given by the covariance between the two variables ($cov(x,y)$) and the noise is expressed by the product between the standard deviations of the two variables ($S_x  S_y$; see Formula\ref{eq:correlation}). Following the conventional standards, in this contribution we will use the symbol $\rho$ to indicate the correlation in the population and the symbol \emph{r} to indicate the value measured in a sample.
\begin{equation}
\label{eq:correlation}
r=\frac{cov(x,y)}{S_x S_y}.
\end{equation}

@@

The central and possibly most difficult point in this process is reasoning on what could be the plausible magnitude and direction of the effect of interest. This critical process, which is central also in traditional a priori power analysis, is an opportunity for researchers to aggregate, formalize and incorporate prior information on the phenomenon under investigation \parencite{gelmanPowerCalculationsAssessing2014}. What is plausible can be determined on theoretical grounds, using expert knowledge elicitations techniques and consulting literature reviews and meta-analysis, always taking into account the presence of effect sizes inflation in the published literature \todo{(Anderson, 2019)}. The plausible effect size approximates the true effect, which is never known but can be thought as “that which would be observed in a hypothetical infinitely large sample” \parencite[p. 642]{gelmanPowerCalculationsAssessing2014}. For a more exhaustive description on plausible effect size, we refer the interested reader to \textcite{altoeEnhancingStatisticalInference2020, gelmanPowerCalculationsAssessing2014}.

@@

To familiarize the reader with Type M and Type S errors, we start our discussion with a retrospective design analysis of a published study. However, the ideal temporal sequence in the research process would be to perform a prospective design analysis in the planning stage of a research project. This is the time when the design is being laid out and useful improvements can be made to obtain more robust results. In this contribution, the order of presentation aims first, to provide an understanding of how to interpret Type M and Type S errors, and then discuss how they could be minimized. The following case study was chosen for illustrative purposes only and, by no means our objective is to judge the study beyond illustrating an application of how to calculate Type M and Type S errors on a published study.

@@

Another important choice in study design is the directionality of the test (i.e., one-tailed or two-tailed).  Design analysis invites reasoning on the plausible effect size and hypothesizing the direction of the effect, not only its magnitude. So why should a researcher perform nondirectional statistical tests when there is a hypothesized direction? Performing a two-tailed test leaves open the possibility to find an unexpected result in the opposite direction \parencite{cohenStatisticalPowerAnalysis1988}, a possibility which may be of special interest for preliminary exploratory studies. However, in more advanced stages of a research program (i.e., confirmatory study), directional hypotheses benefit from higher statistical power and lower Type M error rates (Figure~\ref{fig:Plot_hypothesis}). As an example, let us consider the differences between a two-tailed test and a one-tailed test in the previous case study. We can perform a new prospective design analysis (Figure~\ref{fig:pro_r2}) with plausible correlation of $\rho=.25$, 80\% statistical power, but this time setting the argument \texttt{alternative} in the R function to \texttt{“greater”}. A comparison of the two prospective design analyses, \todo{Figure~\ref{fig:pro_r} and Figure~\ref{fig:pro_r2}}, suggests that the same Type M error rate of about 10\% is guaranteed with 94 participants, instead of 125 subjects. Note that, Type S error is not possible in directional statistical tests. Indeed, all the statistically significant results are obtainable only in the hypothesized direction, not the opposite one.

@@

Peculiar to design analysis is the focus on the implications of design choices on effect sizes estimation rather than statistical significance only. We illustrated how Type M and Type S errors can be prevented with a \emph{prospective design analysis}. In the planning stage of a research project, design analysis has the potential to increase researchers’ awareness of the consequences that their sample size choices have on uncertainty about final estimates of the effects. This favors reasoning in similar terms to those in which results will be evaluated, that is to say, effect size estimation. But understanding the inferential risks in a study design is also beneficial once results are obtained. We presented \emph{retrospective design analysis} on a published study, and the same process can be useful for studies in general, especially those ending without the necessary sample size to maximize statistical power and minimize Type M and Type S errors. In all cases, presenting their values, effectively communicates the uncertainty of the results. In particular, Type M and Type S errors put a red flag when results are statistically significant, but the effect size could be largely overestimated and in the wrong direction. Finally, both prospective and retrospective design analysis favors cumulative science encouraging the incorporation of expert knowledge in the definition of the plausible effect sizes.

>>>>

Effect size exaggeration is a corollary consequence of low statistical power that might not be evident at first. Indeed, statistical power usually focuses on statistical significance, not effect size estimation (i.e., statistical power is the probability to find a statistically significant result if an effect of a certain size actually exists). However, effect size estimation and statistical significance are closely related. This point can be highlighted estimating the Type M (magnitude) and Type S (sign) errors characterizing a study design \parencite{gelmanPowerCalculationsAssessing2014}. Given a study design (i.e., sample size, statistical test directionality, $\alpha$ level and plausible effect size formalization), Type M error indicates the factor by which a statistically significant effect would be, on average, exaggerated. Type S error indicates the probability to find a statistically significant effect in the opposite direction to the one considered plausible. It is evident that both errors are defined starting from a reasoned guess on the plausible magnitude and direction of the effect under study, which is called \emph{plausible effect size} \parencite{gelmanPowerCalculationsAssessing2014}. The analysis that researchers perform to consider Type M and Type S errors in their research practice is called \emph{design analysis}, given the special focus posed into considering the design of a study \parencite{altoeEnhancingStatisticalInference2020, gelmanPowerCalculationsAssessing2014}.

@@

Type M (magnitude) and Type S (sign) errors are not widely known in the psychological research community but their consideration during the research process has the potential to improve current research practices, for example, by increasing the awarness that design choices have on possible studies' results. In a previous work we illustrated Type M and Type S errors using Cohens’\emph{d} as a measure of effect size \parencite{altoeEnhancingStatisticalInference2020}. In this contribution we aim to to further increase familiarity with Type M and Type S errors, considering another common effect size measures in psychology: Pearson correlation coefficient, $\rho$.  The rest of this article is organized as follows: introduction to Type M and Type S errors; description of what is a design analysis and how to conduct one; analysis of Type S and Type M errors when varying alpha levels and hypothesis directionality. Moreover, the two appendices present further implications of design analysis for Pearson correlation (Appendix \ref{sec:appendix_a}) and an extensive illustration of the R functions for design analysis for Pearson correlation (Appendix \ref{sec:appendix_b}).

@@

Pearson correlation coefficient is a standardized effect size measure indicating the strength and the direction of the relationship between two continuous variables \parencite{cohenStatisticalPowerAnalysis1988, ellisEssentialGuideEffect2010}. Even though the correlation coefficient is widely known, we briefly go over its main features using an example. Imagine that we were interested to measure the relationship between anxiety and depression in a population and we plan a study with \emph{n} participants, where, for each participant, we measure the level of anxiety (i.e., variable X) and the level of depression (i.e., variable Y). At the end of the study we will have n pairs of values X and Y. The correlation coefficient helps us answer the questions: how strong is the linear relationship between anxiety and depression in this population? Is the relationship positive or negative? Correlation ranges from -1 to +1, indicating respectively two extreme scenarios of perfect negative relationship and perfect positive relationship\footnote{Correlation indicates a relationship between variables but does not imply causation. We do not discuss
this relevant aspect here but we refer the interested reader to \parencite{rohrerThinkingClearlyCorrelations2018}}. Since the correlation coefficient is a dimensionless number, it is a signal to noise ratio where the signal is given by the covariance between the two variables ($cov(x,y)$) and the noise is expressed by the product between the standard deviations of the two variables ($S_x  S_y$; see Formula\ref{eq:correlation}). Following the conventional standards, in this contribution we will use the symbol $\rho$ to indicate the correlation in the population and the symbol \emph{r} to indicate the value measured in a sample.
\begin{equation}
\label{eq:correlation}
r=\frac{cov(x,y)}{S_x S_y}.
\end{equation}

@@

The central and possibly most difficult point in this process is reasoning on what could be the plausible magnitude and direction of the effect of interest. This critical process, which is central also in traditional power analysis, is an opportunity for researchers to aggregate, formalize and incorporate prior information on the phenomenon under investigation \parencite{gelmanPowerCalculationsAssessing2014}. What is plausible can be determined on theoretical grounds, using expert knowledge elicitations techniques and consulting literature reviews and meta-analysis, always taking into account the presence of effect sizes inflation in the published literature \todo{(Anderson, 2019)}. The plausible effect size approximates the true effect, which is never known but can be thought as “that which would be observed in a hypothetical infinitely large sample” \parencite[p. 642]{gelmanPowerCalculationsAssessing2014}. For a more exhaustive description on plausible effect size, we refer the interested reader to \textcite{altoeEnhancingStatisticalInference2020, gelmanPowerCalculationsAssessing2014}.

@@

To familiarize the reader with Type M and Type S errors, we start our discussion with a retrospective design analysis of a published study. However, the ideal temporal sequence in the research process would be to perform a prospective design analysis in the planning stage of a research project. This is the time when the design is being laid out and useful improvements can be made to obtain more robust results. In this contribution, the order of presentation aims first, to provide an understanding of how to interpret Type M and Type S errors, and then discuss how they could be taken into account. The following case study was chosen for illustrative purposes only and, by no means our objective is to judge the study beyond illustrating an application of how to calculate Type M and Type S errors on a published study.

@@

Another important choice in study design is the directionality of the test (i.e., one-tailed or two-tailed).  Design analysis invites reasoning on the plausible effect size and hypothesizing the direction of the effect, not only its magnitude. So why should a researcher perform nondirectional statistical tests when there is a hypothesized direction? Performing a two-tailed test leaves open the possibility to find an unexpected result in the opposite direction \parencite{cohenStatisticalPowerAnalysis1988}, a possibility which may be of special interest for preliminary exploratory studies. However, in more advanced stages of a research program (i.e., confirmatory study), directional hypotheses benefit from higher statistical power and lower Type M error rates (Figure~\ref{fig:Plot_hypothesis}). As an example, let us consider the differences between a two-tailed test and a one-tailed test in the previous case study. We can perform a new prospective design analysis (Figure~\ref{fig:pro_r2}) with plausible correlation of $\rho=.25$, 80\% statistical power, but this time setting the argument \texttt{alternative} in the R function to \texttt{“greater”}. A comparison of the two prospective design analyses, \todo{Figure~\ref{fig:pro_r} and Figure~\ref{fig:pro_r2}}, suggests that the same Type M error rate of about 10\% is guaranteed with 94 participants, instead of the 125 subjects necessary with a two-tailed test. Note that, Type S error is not possible in directional statistical tests. Indeed, all the statistically significant results are obtainable only in the hypothesized direction, not the opposite one.

@@

Peculiar to design analysis is the focus on the implications of design choices on effect sizes estimation rather than statistical significance only. We illustrated how Type M and Type S errors can be taken into account with a \emph{prospective design analysis}. In the planning stage of a research project, design analysis has the potential to increase researchers’ awareness of the consequences that their sample size choices have on uncertainty about final estimates of the effects. This favors reasoning in similar terms to those in which results will be evaluated, that is to say, effect size estimation. But understanding the inferential risks in a study design is also beneficial once results are obtained. We presented \emph{retrospective design analysis} on a published study, and the same process can be useful for studies in general, especially those ending without the necessary sample size to maximize statistical power and minimize Type M and Type S errors. In all cases, presenting their values, effectively communicates the uncertainty of the results. In particular, Type M and Type S errors put a red flag when results are statistically significant, but the effect size could be largely overestimated and in the wrong direction. Finally, both prospective and retrospective design analysis favors cumulative science encouraging the incorporation of expert knowledge in the definition of the plausible effect sizes.
